---
title: "Workshop on Data analysis with R for Lao PDR’s fourth National Forest Inventory cycle"
subtitle: "Session 05: data preparation"
date: "01-05 Sep 2025"
number-sections: true
toc: true
toc-depth: 2
highlight-style: github
theme: 
  light: flatly
  dark: darkly
format: 
  html: 
    toc-location: left
    embed-resources: true
  docx:
    reference-doc: custom-reference-doc.docx
  pdf: 
    documentclass: scrartcl
    papersize: A4
    colorlinks: true
editor: visual
---

# Goals {.unnumbered}

1.  Load initial tables

2.  Correct errors in subplot table (duplicates in subplot_id)

# Setup {.unnumbered}

Load Packages

```{r}
#| output: false
 
## Useful when sourcing script from Quarto doc.
if (!require(here)) install.packages("here"); library(here)

## Load packages
source(here("R/00-load-packages.R"))
```

<!-- Section 1 -->

# Load harmonized tables and ancillary data

<!-- Section 1.1 -->

## Load subplot, lcs, tree tables from NFI

We can read tables or CSV files from the computer with the function `read_csv()`. We can store the initial table in objects with the suffix '\_init' to keep the initial version in the environment.

Example: read the table training_subplot.csv from the computer into the object `subplot_init`:

```{r}
subplot_init <- read_csv("data/training_subplot.csv", show_col_types = FALSE)
```

Load the tables training_tree.csv and training_lcs.csv in the next exercise (@exr-01).

:::: {.callout-tip appearance="simple"}
<!-- EX-01 -->

::: {#exr-01}
## Load 'training_tree.csv' and 'training_lcs.csv' tables

\linebreak

-   Load the table 'training_lcs.csv' in the R objects 'lcs_init'.
-   Load the table 'training_tree.csv' in the R objects 'tree_init'.

```{r}
#| label: ex-01-code

## Type here the answer to exercise 1

## !!! SOL
lcs_init <- read_csv("data/training_lcs.csv", show_col_types = FALSE)
tree_init <- read_csv("data/training_tree.csv", show_col_types = FALSE)
## !!!
```
:::
::::

<!-- section 1.2 -->

## Load ancillary data (Chave E and Phase 1 plot data)

Ancillary data are loaded into a list to keep the environment tidy.

```{r}
anci <- list()

anci$ph1    <- read_csv("data/training_anci_phase1.csv", show_col_types = FALSE)
anci$plot_E <- read_csv("data/training_anci_plotE.csv", show_col_types = FALSE)
```

<!-- Section 1.3 -->

## Visualize tree locations

We visualize tree locations from the `tree_init` table with `ggplot()` and `geom_point()`.

Step-by-step for one subplot:

-   Define `circ16` as tibble of 100 rows to draw a circular plot in ggplot, based on a radius and an angle "theta" $\theta$ in radian.

-   Draw a ggplot with trees from the table `tree_init` in plot 631 and subplot "c", using `filter()`, and add plot boundary with `circ16`.

```{r}
circ16 <- tibble(
  theta = seq(0, 2*pi, length = 100),
  x = 16 * cos(theta),
  y = 16 * sin(theta)
)

tree_init |>
  filter(plot_no == 631, subplot_no == "C") |>
  ggplot(aes(x = tree_x, y = tree_y)) +
  geom_point() +
  geom_path(data = circ16, aes(x = x, y = y)) +
  coord_fixed()
```

-   In similar way, perform exercise to make a tree location figure for subplot "A" of plot_no 1, but this time apply DBH \<= 30cm and trees with DBH \> 30cm for dbh classes "small" and "big". Also create tibble of circ08 in similar way to add 8m radius circle to the figure

    ```{r}
    ## EX-2: make a tree location figure ####
    ## - for subplot '1A' -> filter()
    ## - change color for trees with DBH <= 30cm and trees with DBH > 30cm -> mutate(), if_else()
    ## - create table circ08 with the data to add the 8m radius circle to the figure
    ## 

    ## !!! SOL
    circ08 <- tibble(
      theta = seq(0, 2*pi, length = 100),
      x = 8 * cos(theta),
      y = 8 * sin(theta)
    )

    tree_init |> 
      filter(plot_no == 1, subplot_no == "A") |>
      mutate(tree_dbh_cat = if_else(tree_dbh <= 30, "small", "big")) |>
      ggplot(aes(x = tree_x, y = tree_y)) +
      geom_point(aes(color = tree_dbh_cat)) +
      geom_path(data = circ08, aes(x = x, y = y)) +
      geom_path(data = circ16, aes(x = x, y = y)) +
      coord_fixed()
    ## !!!
    ```

# Detecting outliers

Here we check for all trees if any of the trees with tree category "small", fall outside the 8m circle. To do this first we need to make a tree map for all small trees and draw 8m and 16m circle on the map to visually observe if any trees fall outside 8m radius.

-   Make a tree location map

```{r}
tree_init |> 
  mutate(tree_dbh_cat = if_else(tree_dbh <= 30, "small", "big")) |>
  filter(tree_dbh_cat == "small") |>
  ggplot(aes(x = tree_x, y = tree_y)) +
  geom_point(aes(color = tree_dbh_cat)) +
  geom_path(data = circ08, aes(x = x, y = y)) +
  geom_path(data = circ16, aes(x = x, y = y)) +
  coord_fixed()
```

-   Observe the tree location map showing "small" trees that fall outside 8m radius

-   Now to identify the trees as outliers

```{r}
## alternative solution
outliers <- tree_init |>
  mutate(tree_dbh_cat = if_else(tree_dbh <= 30, "small", "big")) |>
  filter(tree_dbh_cat == "small", tree_distance > 8)
```

-   Note that 72 observations are identified as outliers

-   Now visualize only the outliers using ggplot by showing points along with8m and 16m radius circles. Include and observe other visualization parameters eg: transparency and symbolics.

    ```{r}
    tree_init |> 
      mutate(tree_dbh_cat = if_else(tree_dbh <= 30, "small", "big")) |>
      filter(tree_dbh_cat == "small") |>
      ggplot(aes(x = tree_x, y = tree_y)) +
      geom_point(alpha = 0.2) +
      geom_point(data = outliers, col = "red", shape = 23, size = 4) +
      geom_path(data = circ08, aes(x = x, y = y)) +
      geom_path(data = circ16, aes(x = x, y = y)) +
      coord_fixed()
    ```

-   Now instead of using tree_dbh\<=30, recode the above script for tree_dbh \<30. This is to check if the 72 observations are actually the outliers or not.

    ```{r}
    ## recoded for DBH < 30 
    tree_init |> 
      mutate(tree_dbh_cat = if_else(tree_dbh < 30, "small", "big")) |>
      filter(tree_dbh_cat == "small") |>
      ggplot(aes(x = tree_x, y = tree_y)) +
      geom_point(aes(color = tree_dbh_cat)) +
      geom_path(data = circ08, aes(x = x, y = y)) +
      geom_path(data = circ16, aes(x = x, y = y)) +
      coord_fixed()
    ## !!!
    ```

-   Observe the ggplot now that after taking tree_dbh \< 30, there are no outliers. It could be due to adoption of different class threshol during calculation. This issue to be discussed in plenary.

# Identify duplicates in subplots

Here we want to check within subplots if there are duplicate ids(subplot_id) and plot numbers(plot_no).

-   Define a vector "vec-dup" to check duplicate subplot ids using group by, summarize, filter and pull functions. Notice the filter and pull functions applied in count \> 1 for identifying more than 1 subplot_id.

    ```{r}
    ## 2.1. Identify duplicates in subplot_id ####
    vec_dup <- subplot_init |>
      group_by(subplot_id) |>
      summarise(count = n(), .groups = "drop") |>
      filter(count > 1) |>
      pull(subplot_id)
    vec_dup
    ```

-   Notice the results as 3 subplots have duplicate subplot_id.

-   Similarly define "vec_dup2" to check duplicate plot_no within each subplots. Here notice that each subplot has 4 plot_nos with same ID, hence the duplicate check should be applied for count \>4.

    ```{r}
    ## EX: Identify duplicates of subplot_plot_no ####
    ## !!! SOL
    vec_dup2 <- subplot_init |>
      group_by(plot_no) |>
      summarise(count = n(), .groups = "drop") |>
      filter(count > 4) |>
      pull(plot_no)
    vec_dup2
    ## !!!
    ```

-   Now to correct the duplicate subplot_id, first check each of the duplicate subplots by filtering them individually and observing the time-stamps of data collection to decide on which subplot_id need correction.

    ```{r}
    ## 2.2. Correct the subplot_no issues ####

    ## How to correct: check the duplicates and observe the timestamps
    tt <- subplot_init |> filter(plot_no == 631)
    ```

-   Perform the correction using case_when function for subplot_id 631C

    ```{r}
    subplot <- subplot_init |>
      mutate(
        subplot_no = case_when(
          subplot_id == "631C" & ONA_index == 109 ~ "B",
          ## ADD more corrections,
          TRUE ~ subplot_no
        )
      )
    ```

<!-- -->

-   Prform the same correction using case_when function for all subplot_id

```{r}
subplot <- subplot_init |>
  mutate(
    subplot_no = case_when(
      subplot_id == "631C" & ONA_index == 109 ~ "B",
      subplot_id == "632C" & ONA_index == 113 ~ "B",
      subplot_id == "553D" & ONA_index == 265 ~ "C",
      TRUE ~ subplot_no
    )
)
```

-   In addition, since subplot_id is a text field, we need to make it consistent for number of characters to ensure correct sorting of the rows. For this purpose perform additional correction in subplot_id to include texts "0" and "00" to make the number of characters consistent.

```{r}
subplot <- subplot_init |>
  mutate(
    subplot_id = case_when(
      plot_no < 10 ~ paste0("00", plot_no, subplot_no),
      plot_no < 100 ~ paste0("0", plot_no, subplot_no),
      TRUE ~ paste0(plot_no, subplot_no)
    )
  )
```

-   Finally check duplicates again to ensure corrections are applied

    ```{r}
    vec_dup <- subplot |>
      group_by(subplot_id) |>
      summarise(count = n(), .groups = "drop") |>
      filter(count > 1) |>
      pull(subplot_id)
    vec_dup
    ```

# Checks in tree and landcover tables

Perform checks in tree related to tree_dbh, azimuth and distance to ensure no absurd values are present.

```{r}
## Check that the min DBH is bigger than 10 and that the biggest DBH is realistiic with summary()
summary(tree_init$tree_dbh)

## EX: check azimuth and distance ####
## - check that azimuth is between 0 and 360
## - check that distance is between 0 and 16

## !!! SOL
summary(tree_init$tree_azimuth)
summary(tree_init$tree_distance)
## !!!
```

-   After resolving all errors, save as main tables for tree and lcs

    ```{r}
    ## >> no errors saving to main
    tree <- tree_init
    lcs <- lcs_init


    ## Remove temporary objects
    rm(vec_dup, vec_dup2, tt)
    ```
