---
title: "Workshop on Data analysis with R for Lao PDR’s fourth National Forest Inventory cycle"
subtitle: "Session 05: data preparation"
date: "01-05 Sep 2025"
number-sections: true
toc: true
toc-depth: 2
highlight-style: github
theme: 
  light: flatly
  dark: darkly
format: 
  html: 
    toc-location: left
    embed-resources: true
  docx:
    reference-doc: custom-reference-doc.docx
  pdf: 
    documentclass: scrartcl
    papersize: A4
    colorlinks: true
editor: visual
---

# Goals {.unnumbered}

1.  Load initial tables and visualize core data

2.  Correct errors in subplot table (duplicates in subplot_id)

# Setup {.unnumbered}

Load packages by sourcing a script with `source()`.

```{r}
#| output: false
 
## Useful when sourcing script from Quarto doc.
if (!require(here)) install.packages("here"); library(here)

## Load packages
source(here("R/00-load-packages.R"))
```

<!-- Section 1 -->

# Load harmonized tables and ancillary data

<!-- Section 1.1 -->

## Load subplot, lcs, tree tables from NFI

We can read tables or CSV files from the computer with the function `read_csv()`. We can store the initial table in objects with the suffix '\_init' to keep the initial version in the environment.

Example: read the table 'training_subplot.csv' from the computer into the object `subplot_init`:

```{r}
subplot_init <- read_csv("data/training_subplot.csv", show_col_types = FALSE)
```

Load the tables 'training_tree.csv' and 'training_lcs.csv' in the next exercise (@exr-01).

:::: {.callout-tip appearance="simple"}
<!-- EX-01 -->

::: {#exr-01}
## Load 'training_tree.csv' and 'training_lcs.csv' tables

\linebreak

-   Load the table 'training_lcs.csv' in the R objects 'lcs_init'.
-   Load the table 'training_tree.csv' in the R objects 'tree_init'.

```{r}
#| label: ex-01-code

## Type here the answer to exercise 1

```
:::
::::

<!-- section 1.2 -->

## Load ancillary data

**Ancillary data are not collected in the field but useful for the analysis of field data.**

In the training we use plot level environment factor (E) from Chave et al. 2014 and the NFI Phase 1 plot data. Ancillary data are loaded into a list to keep the environment tidy.

```{r}
anci <- list()

anci$ph1    <- read_csv("data/training_anci_phase1.csv", show_col_types = FALSE)
anci$plot_E <- read_csv("data/training_anci_plotE.csv", show_col_types = FALSE)
```

<!-- Section 1.3 -->

## Visualize tree locations

We visualize tree locations from the `tree_init` table with `ggplot()` and `geom_point()` to see if any potential error.

Step-by-step for one subplot:

-   Define `circ16` as tibble of 100 rows to draw a circular plot in ggplot, based on a radius and an angle "theta" $\theta$ in radian.

-   Draw a ggplot with trees from the table `tree_init` in plot 631 and subplot "C", using `filter()`,

-   Use `tree_x` and `tree_y` to show trees based on their location from the subplot center,

-   Add plot boundary with `circ16`.

```{r}
circ16 <- tibble(
  theta = seq(0, 2*pi, length = 100),
  x = 16 * cos(theta),
  y = 16 * sin(theta)
)

tree_init |>
  filter(plot_no == 631, subplot_no == "C") |>
  ggplot(aes(x = tree_x, y = tree_y)) +
  geom_point() +
  geom_path(data = circ16, aes(x = x, y = y)) +
  coord_fixed()
```

Adapt the previous code to make a tree location map for the subplot '1A' and add different colors based on tree DBH size in @exr-02.

:::: {.callout-tip appearance="simple"}
<!-- EX-02 -->

::: {#exr-02}
## Make a tree location figure

\linebreak

-   create the table `circ08` similar to `circ16` but for a 8m radius circle.

-   Filter within `tree` the trees from subplot '1A' with `filter()`.

-   Use `mutate()` and `if_else()` to add a column `tree_dbh_cat` that has value "small" for trees with DBH \< 30 cm and "big" for the other trees.

-   Make the figure of tree location with the 8m and 16m radius circles and change the color of trees based on `tree_dbh_cat`.

```{r}
#| label: ex-02-code

## Type here the answer to exercise 2


```
:::
::::

Visualizations can also be used to detect outliers. We use the example of small trees outside the nested subplot circle of 8m radius as guided exercise in @exr-03.

:::: callout-tip
<!-- EX-03 -->

::: {#exr-03}
## (Optional) detect outliers on graph

\linebreak

First, make a figure with tree location of all trees smaller than or equal to 30 cm with the 2 circles of 8 and 16 meters.

-   use filter() to get all trees with DBH \<= 30 cm.

-   make the ggplot of tree locations

```{r}
#| label: ex-03-1

## Type here the answer to exercise 3 part 1


```

We can highlight outliers by making a specific table for them, they are trees with `tree_distance` \> 8m and `tree_dbh` \<= 30 cm. We can plot that table with big size point and a shape with no fill (see shape 21 to 24 in `help(pch)`.

Make a new figure with:

-   Create the table `outliers` containing trees with DBH \<= 30 and distance \> 8. Use `filter()`.

-   Make the same figure as before and add a `geom_point()` with `data = outliers`, `shape = 23` and `size = 4` .

```{r}
#| label: ex-03-2

## Type here the answer to exercise 3 part 2


```
:::

TIP: here we added `alpha = 0.2` to the geometry of the trees. It make tree point semi-transparent and helps to detect potential measurement issues if we would detect patches with few trees.

Now are these outliers measurement errors?

-   Make the same graph as this exercise's part 1, but this time with tree DBH \< 30 and [**not**]{.underline} DBH \<= 30.

```{r}
#| label: ex-03-3

## Type here the answer to exercise 3 part 3


```

Conclusion: the trees found outside the small circle were trees with DBH = 30 cm. They were not measurement errors, they were part of the big trees and therefore could be found outside the 8m radius circle.
::::

<!-- Section 2 -->

# Find and correct errors

One of the main issue found in the raw data was the entry errors of subplot and tree IDs. It is very important in Database Management Systems (DBMS) to have unique IDs for all tables. In NFI it is particularly important for joining information from one table to the other. For example, land cover information is recorded at subplot level and needs to be passed on to tree level to apply the correct allometric equations. These errors can be detected programatically but often have to be solved manually.

This section introduces how to use `group_by()` and `summarise()` to find duplicates in subplot IDs

<!-- section 2.1 -->

## Identify duplicates in subplot IDs

Here we want to check within subplots if there are duplicates in subplot IDs (`subplot_id`) and plot numbers (`plot_no`) in the `subplot_init` table. The objective is to create a vector of subplot Ids that are duplicated. If there are no duplicate the vector will be of length 0.

step-by-step:

-   Create a vector `vec_dup` and assign it the following code sequence.

-   Group the subplot table by subplot ID with `group_by()` and use `summarise()` to count the number of subplots that have each ID inside a new column `count`. Use the function `n()` to count.

-   Filter all the subplot IDs that have a count bigger than 1 with `filter()`.

-   Pull the subplot IDs.

```{r}
vec_dup <- subplot_init |>
  group_by(subplot_id) |>
  summarise(count = n(), .groups = "drop") |>
  filter(count > 1) |>
  pull(subplot_id)
vec_dup
```

Notice the results as 3 subplots have duplicate `subplot_id`.

Now practice the same code sequence to check if there are duplicates in `plot_no` in the @exr-04. Note that each plot ID should appears 4 times in the subplot table, one for each subplot, so duplicates are when `count > 4`.

:::: {.callout-tip appearance="simple"}
<!-- EX-04 -->

::: {#exr-04}
## Identify duplicates of plot number

\linebreak

-   Create `vec_dup2` similarly to `vec_dup`, but this time looking for plot_no that have more than 4 subplot IDs, so you will need to group by `plot_no` and filter `count > 4` and pull `plot_no`.

```{r}
#| label: ex-04

## Type here the answer to exercise 4


```

There should be no duplicates of `plot_no`.
:::
::::

## Correct the subplot ID issues

To correct the duplicates in subplot_id, we filter the subplots from one of the duplicated IDs and visually check what is wrong. In case one ID is missing and one is duplicated we use timestamps to identify which of the duplicates should be renamed.

Then we can use `mutate()` and `case_when()` to apply correction in R.

Step-by-step:

-   Filter the subplots of one plot with duplicates: 631C.

```{r}
tt <- subplot_init |> filter(plot_no == 631)
```

-   Visualize the table (run `View(tt)` or click `tt` in the Environment tab (top-right window in Rstudio).

    -   In the case of '631C' there are 2 subplots C but no B. we can use timestamps to identify that `ONA_index` 109 should actually be for subplot B.

-   Perform the correction using case_when() function for subplot_id 631C. At this stage only ONA_index is unique so it can be used to correctly identify the subplot to correct.

```{r}
subplot <- subplot_init |>
  mutate(
    subplot_no = case_when(
      subplot_id == "631C" & ONA_index == 109 ~ "B",
      ## ADD more corrections,
      TRUE ~ subplot_no
    )
  )
```

You can now copy/paste the above R chunk and complete the correction for the 2 other subplot IDs with duplicates in @exr-05.

Note: we correct `subplot_no` now, but we will need to remake `subplot_id` once `subplot_no` is correct.

:::: {.callout-tip appearance="simple"}
<!-- EX-05 -->

::: {#exr-05}
## Correct duplicates in subplot IDs

\linebreak

-   for each remaining duplicate, run the table `tt` with all subplots for the plot with duplicates

-   Visualize `tt` and identify `ONA_index` of the subplot that should be corrected and what is the correct `subplot_no`.

-   Copyu/paste the `subplot <- ...` sequence from above and fill in the `case_when()` with all the correction for the 3 duplicated subplots.

```{r}
#| label: ex-05

## Type here the answer to exercise 5


```
:::
::::

Now we can recreate `subplot_id` from trhe corrected `subplot_no`. Since `subplot_id` is a text field, we need to make it consistent for number of characters to ensure correct sorting of the rows. For this purpose we include texts "0" and "00" to make the number of characters consistent.

```{r}
subplot <- subplot_init |>
  mutate(
    subplot_id = case_when(
      plot_no < 10 ~ paste0("00", plot_no, subplot_no),
      plot_no < 100 ~ paste0("0", plot_no, subplot_no),
      TRUE ~ paste0(plot_no, subplot_no)
    )
  )
```

Now if we re-run the `vec_dup` sequence with `subplot` instead of `subplot_init`, `vec_dup` should be of length 0.

```{r}
vec_dup <- subplot |>
  group_by(subplot_id) |>
  summarise(count = n(), .groups = "drop") |>
  filter(count > 1) |>
  pull(subplot_id)
vec_dup
```

## Check outliers in tree table

The tree table we can check for outliers in numeric columns with the `summary()` function.

For example with DBH:

```{r}
summary(tree_init$tree_dbh)
```

There are no trees with DBH \< 10 cm or unrealistically big DBH, all good.

We can check for outliers in tree distance and azimuth in @exr-06.

:::: {.callout-tip appearance="simple"}
<!-- EX-06 -->

::: {#exr-06}
## Check azimuth and distance

\linebreak

-   Check that azimuth is between 0 and 360

-   Check that distance is between 0 and 16

```{r}
#| label: ex-06

## Type here the answer to exercise 6

```

No error found in distance or azimuth.
:::
::::

After resolving all errors, save the tables to the entity name without "\_init" suffix. This way we have both the initial and corrected version in the environment. If tehre were no errors the inital tables can be passed directly to the corrected tables:

```{r}
tree <- tree_init
lcs <- lcs_init

```

Finally we can remove the temporary objects that won't be used later.

```{r}
rm(vec_dup, vec_dup2, tt, outliers)
```

NOTE: In the main analysis the temporary objects are often stored in a list called `tmp`. This avoids cluttering the environment with temporary objects and makes it easy to remove them with `rm(tmp)`.

```{r}
#| eval: false

## Create empty list
tmp <- list()

## Populate the list (example)
tmp$outliers <- tree_init |> filter(tree_dbh <= 30, tree_distance > 8)

## Remove temporary objects
rm(tmp)

```
