---
title: "Workshop on Data analysis with R for Lao PDR’s fourth National Forest Inventory cycle"
subtitle: "Session 07: tree weight and basal area"
date: "01-05 Sep 2025"
number-sections: true
toc: true
toc-depth: 2
highlight-style: github
theme: 
  light: flatly
  dark: darkly
format: 
  html: 
    toc-location: left
    embed-resources: true
  docx:
    reference-doc: custom-reference-doc.docx
  pdf: 
    documentclass: scrartcl
    papersize: A4
    colorlinks: true
editor: visual
---

# Goals {.unnumbered}

1.  Calculate tree weights for ratio estimators and for subplot per ha values

2.  Calculate tree basal area

# Setup {.unnumbered}

```{r}
#| label: setup
#| output: false
 
## Useful when sourcing script from Quarto doc.
if (!require(here)) install.packages("here"); library(here)

source(here("R/00-load-packages.R"))
source(here("R/SOL-05-data-preparation.R"))
source(here("R/SOL-06-data-joins.R"))
```

<!-- Section 1 -->

# Tree weights for nested circles adjustment

<!-- Section 1.1 -->

## Tree weight for ratio estimator

In the context of **ratio estimator,** tree weight converts small size trees' variables to their equivalent if small trees were measured in the same area as the bigger sized trees.

For Lao PDR, the ratio between small and big trees is 4:

```{r}
#| label: weight_re

tree <- tree |> mutate(tree_weight = if_else(tree_dbh < 30, 4, 1))

```

Let's calculate where this value 4 comes from in @exr-71.

:::: {.callout-warning appearance="simple"}
<!-- Ex 7.1 -->

::: {#exr-71}
## Where does weight 4 come from?

\linebreak

Instructions:

-   Calculate 'Asmall' as the area of a circle of 8m radius

-   Calculate 'Abig' as the area of a circle of 16m radius

-   Calculate Abig / Asmall
:::
::::

Type here answers to @exr-71:

```{r}
#| label: ex-71

## Your code here

##!!! SOL
Asmall <- pi * 8^2 
Abig   <- pi * 16^2
Abig / Asmall 
## !!!
```

<!-- Section 1.2 -->

## Tree weight for subplot per ha values

To get per ha at the subplot level, we first assign to each tree its subplot nested circle area, then the weight is the inverse of this area

```{r}
#| label: weight_sperr

tree <- tree |> 
  mutate( 
    tree_spha = if_else(tree_dbh < 30, pi * 16^2 / 10000, pi * 16^2 / 10000),
    tree_weight_spha = 1 / tree_spha
  )

```

There might be an error in the above code snippet, let's solve it in @exr-72.

:::: {.callout-warning appearance="simple"}
<!-- EX 7.2 -->

::: {#exr-72}
## Find the error

\linebreak

Instructions:

-   Copy/paste the previous code snippet

-   Find and correct the error in `tree_spha`
:::
::::

Type here answers to @exr-72:

```{r}
#| label: ex-72

## Your code here

##!!! SOL
tree <- tree |> mutate( 
  tree_spha = if_else(tree_dbh < 30, pi * 8^2 / 10000, pi * 16^2 / 10000), 
  tree_weight_spha = 1 / tree_spha 
  )
## !!!
```

<!-- Section 2 -->

## Calculate tree basal area

The basal area of a tree is its stem area footprint based on the measurement of it diameter at breast height.

```{r}
#| label: tree-ba

tree <- tree |> mutate(tree_ba = pi * (tree_dbh/200)^2)

```

Let's check subplot level basal area per land cover class in @exr-73.

:::: {.callout-warning appearance="simple"}
<!-- EX 7.3 -->

::: {#exr-73}
## (optional) Basal area per subplot

\linebreak

Instructions:

-   From tree, use `group_by()` and `summarise()` to create `subplot_ba`, the sum of `tree_ba * tree_weight_spha` per plot and subplot number,

-   Join `lc_class_center` from the `subplot` table using suffix control,

-   Make a boxplot of `subplot_ba` against `lc_class_center`.
:::
::::

Type here answers to @exr-73:

```{r}
#| label: ex-73


## Your code here
## !!! SOL
tmp_sp <- subplot |> select(plot_no, subplot_no, subplot_lc_class_center) 

tree |> 
  group_by(plot_no, subplot_no) |> 
  summarise(subplot_ba = sum(tree_ba * tree_weight_spha), .groups = "drop") |> 
  left_join(tmp_sp, by = join_by(plot_no, subplot_no)) |> 
  ggplot(aes(x = subplot_lc_class_center, y = subplot_ba)) + 
  geom_boxplot() + 
  geom_jitter(alpha = 0.1, shape = 21) ## bonus to see the observations on top of the boxplot
## !!!

```
