---
title: "Workshop on Data analysis with R for Lao PDR’s fourth National Forest Inventory cycle"
subtitle: "Session 08: tree aboveground biomass"
date: "01-05 Sep 2025"
number-sections: true
toc: true
toc-depth: 2
highlight-style: github
theme: 
  light: flatly
  dark: darkly
format: 
  html: 
    toc-location: left
    embed-resources: true
  docx:
    reference-doc: custom-reference-doc.docx
  pdf: 
    documentclass: scrartcl
    papersize: A4
    colorlinks: true
editor: visual
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| output: false
 
## Useful when sourcing script from Quarto doc.
if (!require(here)) install.packages("here"); library(here)

source(here("R/00-load-packages.R"))
source(here("R/SOL-05-data-preparation.R"))
source(here("R/SOL-06-data-joins.R"))
source(here("R/SOL-07a-tree-weight.R"))
source(here("R/SOL-07b-tree-basal-area.R"))
```

# Session 08: tree aboveground biomass

**GOALS:**

1.  add AGB at tree level for each forest type

2.  Add AGB from chave 05 and 14 without height

3.  Make a figure for natural forest and one for plantations, showing forest type AGB, and chave models

<!-- Section 8.1 -->

## List of models

-   `"EG" ~ 0.3112 * tree_dbh^2.2331,`

-   `"MD" ~ 0.523081 * tree_dbh^2,`

-   `"DD" ~ 0.2137 * tree_dbh^2.2575,`

-   `"CF" ~ 0.1277 * tree_dbh^2.3944,`

-   `"MCB" ~ 0.1277 * tree_dbh^2.3944,`

-   `"P_AC" ~ 0.1173 * tree_dbh^2.454,`

-   `"P_EC" ~ 0.199 * tree_dbh^2.185,`

-   `"P_RB" ~ 0.0082 * (pi*tree_dbh)^2.5623, ## Rubber model uses circumference`

-   `"P_TK" ~ 0.077 * tree_dbh^2.546,`

-   `"P_OTH" ~ 0.3112 * tree_dbh^2.2331,`

-   `"RV" ~ 0.6 * exp(-1.499 + 2.148 * log(tree_dbh) + 0.207 * (log(tree_dbh))^2 - 0.0281*(log(tree_dbh))^3),`

-   `"B" ~ 0.6 * exp(-1.499 + 2.148 * log(tree_dbh) + 0.207 * (log(tree_dbh))^2 - 0.0281*(log(tree_dbh))^3),`

-   `Chave14 = round(exp(-1.803 - 0.976plot_E + 0.976log(0.6) + 2.673log(tree_dbh) -0.0299(log(tree_dbh))^2)`

-   `Chave05 = round(0.6 * exp(-1.499 + 2.148log(tree_dbh) + 0.207(log(tree_dbh))^2 - 0.0281*(log(tree_dbh))^3)`

<!-- Section 8.2 -->

## AGB per forest type

We can assign different AGB models to trees for different forest types with `mutate()` and `case_when()`:

```{r}
#| label: agb-demo

#table(tree$lc_code)

tree <- tree |> 
  mutate( 
    tree_agb_final = case_when(
      lc_code == "EG" ~ 0.3112 * tree_dbh^2.2331, 
      ## ADD OTHER equations here 
      TRUE ~ 0
    ) 
  )
```

Let's fill in the other cases in the @exr-81.

:::: {.callout-warning appearance="simple"}
<!-- EX 8.1 -->

::: {#exr-81}
## \~ Complete tree_agb_final

\linebreak

-   For all forest land covers add their equations as shown at the beginning of the document.

-   For land cover with no equation, assign 0 with `TRUE ~ 0`.

-   Make a figure with 'tree_agb_final' against 'tree_dbh' as points (optional)

-   Make a figure with 'tree_agb_final' against 'tree_dbh' as line with color based on 'lc_code' (optional)
:::
::::

Type here answers to @exr-81:

```{r}
#| label: ex-81-code

## Your code here

```

<!-- Section 8.3 -->

## AGB from Chave 2005 and 2014

Let's add 3 models to all trees, regardless of their forest type to compare with the forest type based models, in @exr-82.

:::: {.callout-warning appearance="simple"}
<!-- EX 8.2 -->

::: {#exr-82}
## \~ Add AGB allometric equations for all trees

\linebreak

Instructions:

-   Add 'tree_agb_chave05' with Chave et al. 2005 equation

-   Add 'tree_agb_chave14' with Chave et al. 2014 equation

-   Add 'tree_agb_EG' with the evergreen forest model
:::
::::

Type here answers to @exr-82:

```{r}
#| label: ex-82-code

## Your code here

```

<!-- Section 8.4 -->

## Compare AGB models

<!-- Section 8.4.1 -->

### Natural forest

Let's compare our forest type specific tree aboveground with the Chave's models and the model for Evergreen forests. Since there are many forest type let's look at natural forests first with `filter()`.

```{r}
#| label: agb-compa-natural

tree |>
  filter(tree_agb_final != 0, lc_no <= 20) |>
  ggplot(aes(x = tree_dbh, y = tree_agb_final)) +
  geom_point() +
  geom_line(aes(y = tree_agb_chave05), color = "grey60") +
  geom_line(aes(y = tree_agb_chave14), color = "darkred") +
  geom_line(aes(y = tree_agb_EG), color = "darkgreen") +
  facet_wrap(~lc_no)
```

We can observe in this figure that the forest type specific models are quite conservative, since our AGB final model predicts smaller AGB for the same DBH compared to both Chave's models.

<!-- Section 8.4.2 -->

### (optional) Planted forest

Let's produce the same figure for planted forests in @exr-83.

:::: {.callout-warning appearance="simple"}
<!-- EX 8.3 -->

::: {#exr-83}
## \~ Compare models for plantations

\linebreak

Instructions:

-   Make a similar graph than 3.1 but for planted forest only
:::
::::

Type here answers to the @exr-83:

```{r}
#| label: ex-83-code

## Your code here

```

Plantation models selected for AGB final are also conservative in general, except for the rubber model.
