---
title: "Workshop on Data analysis with R for Lao PDR’s fourth National Forest Inventory cycle"
subtitle: "Session 09: Aggregation"
date: "01-05 Sep 2025"
number-sections: true
toc: true
toc-depth: 2
highlight-style: github
theme: 
  light: flatly
  dark: darkly
format: 
  html: 
    toc-location: left
    embed-resources: true
  docx:
    reference-doc: custom-reference-doc.docx
  pdf: 
    documentclass: scrartcl
    papersize: A4
    colorlinks: true
editor: visual
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| output: false
 
## Useful when sourcing script from Quarto doc.
if (!require(here)) install.packages("here"); library(here)

source(here("R/00-load-packages.R"))
source(here("R/fct-nfi-aggregate3-new.R"))

source(here("R/SOL-05-data-preparation.R"))
source(here("R/SOL-06-data-joins.R"))
source(here("R/SOL-07a-tree-weight.R"))
source(here("R/SOL-07b-tree-basal-area.R"))
source(here("R/SOL-08a-tree-agb.R"))
```

# Session 09: Aggregate tree to forest type AGB

**GOALS:**

1.  Calculate the mean aboveground biomass for all forest types using **plot level** land cover and **systematic sampling**.
2.  Calculate the mean aboveground biomass for all forest types using **ratio estimators** and **2-stages sampling for post-stratification**.
3.  Compare estimates.

<!-- Section 9.1 -->

## AGB with plot level land cover and systematic sampling

**Step-by-step**

1.  Sum tree AGB to subplots.

2.  Make plot majority LC class.

3.  Create plot level AGB based on mean subplot AGB.

4.  Create Forest type AGB as the mean of plot AGB per forest type.

5.  Calculate sampling error.

### Sum tree AGB to subplot

We will use `group_by()` and `summarise()` to bring tree characteristics to the subplot level.

```{r}
#| label: sp-agb

subplot_agb <- tree |>
  group_by(plot_no, subplot_no) |>
  summarise(
    sp_agb = sum(tree_agb_final * tree_weight_spha / 1000),
    sp_ba  = sum(tree_ba * tree_weight_spha),
    sp_dens = sum(tree_weight_spha),
    tree_meas = n()
  )
```

### Load plot level main land cover

Plot level land cover was generated, from the subplots center LC, with the following rules:

1.  For plots with one unique LC, they were directly assigned to plot level.
2.  For plots with multiple land cover, we used the majority LC for the whole plot.
3.  If some plots have equal importance of two land covers, we used to minimum land cover class.

```{r}
#| label: load-plotlc

plot_lc <- read_csv("data/training_plot_lc.csv", show_col_types = FALSE)

```

### Aggregate subplot to plot

We take the mean of subplot variables (except tree measured count).

```{r}
#| label: plot-agb

plot_agb <- subplot_agb |>
  group_by(plot_no) |>
  summarise(
    plot_agb = mean(sp_agb),
    plot_ba = mean(sp_ba),
    plot_dens = mean(sp_dens),
    plot_tree_meas = sum(tree_meas)
  ) |>
  left_join(plot_lc, by = join_by(plot_no))

```

Let's look at plot AGB per land cover in @exr-91.

:::: {.callout-warning appearance="simple"}
<!-- EX 9.1 -->

::: {#exr-91}
## \~ Boxplot of plot AGB

\linebreak

Instructions:

-   Make a boxplot of plot AGB against land cover classes
:::
::::

Type here the answers to @exr-91:

```{r}
#| label: ex-91-code

## Your code here
plot_agb |>
  ggplot(aes(x = plot_lc, y = plot_agb)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1, size = 0.8)
```

### Aggregate to forest type AGB

Now we can take averages of plot AGB for each forest type. For the uncertainty, we also need **standard deviation** of AGB and the **number of plots**.

```{r}
#| label: ftype-agb

ftype_agb <- plot_agb |>
  group_by(plot_lc) |>
  summarise(
    plot_count = n(),
    total_tree_meas = sum(plot_tree_meas),
    ftype_dens = mean(plot_dens),
    ftype_ba = mean(plot_ba),
    ftype_agb = mean(plot_agb),
    ftype_agb_sd = sd(plot_agb)
  )

```

We can calculate the uncertainty for systematic design with this formula:

$$
U\% = t^{1 - \alpha/2}_{n-1} \times \frac{\sigma}{\sqrt{n}} \times \frac{1}{\mu} \times 100
$$

```{r}
#| label: ftype-agbu

ftype_agb <- ftype_agb |>
  mutate(
    ftype_agb_se = ftype_agb_sd / sqrt(plot_count),
    ftype_agb_me = qt(0.975, plot_count - 1) * ftype_agb_se,
    ftype_agb_U  = ftype_agb_me / ftype_agb * 100,
    ftype_agb_cilower = ftype_agb - ftype_agb_me,
    ftype_agb_ciupper = ftype_agb + ftype_agb_me
  )

```

Make a plot of forest type AGB with sampling uncertainty

```{r}
#| label: ftype-agb-plot

ftype_agb |>
  ggplot(aes(x = plot_lc, y = ftype_agb)) +
  geom_col(aes(fill = plot_lc), col = "black") +
  geom_errorbar(aes(ymin = ftype_agb_cilower, ymax = ftype_agb_ciupper)) +
  theme(legend.position = "none") +
  ylim(0, 600)

```

<!-- Section 9.2 -->

## AGB with ratio estimator and 2-stages sampling for post-stratification

<!-- Section 9.3 -->

## Compare AGB estimates from the 2 aggregation methods
